---
title: "Data_Analysis_2023_AW"
author: "Arne Weigelt"
date: "2023-05-02"
output: html_document
---


load data
```{r}
#load("C:/Users/ArneW/OneDrive - bwedu/Bioinformatics/DATA_ANALYSIS_2023/prism_datasets.rda")
#load("C:/Users/ArneW/OneDrive - bwedu/Bioinformatics/DATA_ANALYSIS_2023/cellline_datasets.rda")


install.packages("tidyverse", "data.table", "magrittr", "reshape2")
install.packages("BiocManager")
BiocManager::install("sva")
install.packages("reshape2")

library(tidyverse)
library(data.table)
library(magrittr)
library(reshape2)
library(sva)
library(dplyr)

library(pheatmap)
```

first look
```{r}
head(prism)
```
Histogramm der ersten Spalte
```{r}
plot(prism[,1])
hist(prism[,1], breaks=150)
```

NA Werte looki looki
```{r}
colnames(prism)[1]
sum(is.na(prism$"BRD-A00077618-236-07-6::0.00061034::HTS002"))
```

 Berechne/plot den Mittelwert/SD jeder Spalte des Dataframes, wobei NA-Werte ignoriert werden
```{r}
prism.col.mean= apply(prism, 2, function(x){mean(x, na.rm= T)})
plot(prism.col.mean, ylab= "column means", xlab="treatments", main= "prism" )



prism.col.var= apply(prism, 2, function(x){var(x, na.rm= T)})
plot(prism.col.var, ylab= "column variance", xlab="treatments", main= "prism" )

#vektor mit kidney cl names
cl.kidney.names= rownames(cl.kidney)



```
Funktion, welche  columns nach parameter drug extrahiert in noch leeres df
```{r}
extract_columns <- function(prism, drug) {
  
  col_indices <- grep(drug, colnames(prism))
  
  # Erstelle ein neues Dataframe mit ausgewählten Spalten
  new_prism <- prism[, col_indices]
  
  return(new_prism)
}

```

extrahiert drug name // hätte man auch mit prism.treat machen können
`````{r}
prism.col.names= unique(sapply(strsplit(colnames(prism),"::"),`[`, 1))
```

fasst alle Spalten einer drug und Zeile zsm., erstellt mean, neues df
```{r}
prism.drug.means= data.frame(matrix(nrow= nrow(prism), ncol=0))
rownames(prism.drug.means)= rownames(prism)
for(drug in prism.col.names){
  df= extract_columns(prism, drug)
 prism.drug.means[drug]= rowMeans(df, na.rm=T)
}

```


```{r}
#heatmap des mean aller einzelnen treatments für alle cl ohne NaN
pheatmap(na.omit(prism.drug.means), show_colnames=F, show_rownames=F, main= " cell line viability of prism.drug.means (na.omit) ")
```
zeigt alle moa o.ä. der drugs mit einem drug mean unter threshold x an // moa und threshold änderbar
```{r}
idx= colnames(prism.drug.means[which(apply(prism.drug.means, 2, min, na.rm=T)< -1.8)])

unique(prism.treat[prism.treat$broad_id %in% idx,]$moa)
```



```{r}
#Define q10
prism.drug.means.q= quantile(prism.drug.means, c(0.02, 0.05, 0.1, 0.25), na.rm=T )
prism.drug.means.kidney.q= quantile(prism.drug.means.kidney, c(0.02, 0.05, 0.1, 0.25), na.rm=T )
#Histo kidney cancer drug mean wirkung in qX

prism.drug.means.col.mean= colMeans(prism.drug.means, na.rm=T)
plot(prism.drug.means.col.mean, ylab= "column means", xlab="treatment mean", main= "prism.drug.mean")
abline(h= prism.drug.means.q, col=c("darkblue", "lightblue", "orange", "red"))

prism.drug.means.col.var= apply(prism.drug.means, 2, function(x){var(x, na.rm= T)})
plot(prism.drug.means.col.var,  ylab= "column variance", xlab="treatment mean", main= "prism.drug.mean")

prism.drug.means.kidney.col.mean= colMeans(prism.drug.means.kidney, na.rm=T)
plot(prism.drug.means.kidney.col.mean,  ylab= "column means", xlab="treatment mean", main= "prism.drug.mean.kidney")
abline(h=prism.drug.means.kidney.q, col=c("darkblue", "lightblue", "orange", "red"))

prism.drug.means.kidney.col.var= apply(prism.drug.means.kidney, 2, function(x){var(x, na.rm= T)})
plot(prism.drug.means.kidney.col.var,  ylab= "column variance", xlab="treatment mean", main= "prism.drug.mean.kidney")
abline(h=prism.drug.means.kidney.q, col=c("darkblue", "lightblue", "orange", "red"))



#!!NOT WORKING YET!!:
prism.drug.means.q10= prism.drug.means<q10
# Zähle die Anzahl der TRUE-Werte in jeder Spalte (ignoriert NA-Werte)
col_counts = colSums(isTRUE(prism.drug.means.q10))

# Finde Spalten, die nur FALSE enthalten
cols_to_remove = col_counts == 0

# Entferne die entsprechenden Spalten
prism.drug.means.q10 = prism.drug.means.q10[, !cols_to_remove, drop = FALSE]



```



```{r}
#ALLES FÜR ROWS(CL)



prism.drug.means.row.mean= rowMeans(prism.drug.means, na.rm=T)

#Define q10
prism.drug.means.row.q= quantile(prism.drug.means.row.mean, probs= c(0.02, 0.05, 0.1, 0.25), na.rm=T )
prism.drug.means.kidney.row.q= quantile(prism.drug.means.kidney.row.mean, c(0.02, 0.05, 0.1, 0.25), na.rm=T )
#Histo kidney cancer cl mean  in qX

plot(prism.drug.means.row.mean, ylab= "row means", xlab="treatment mean", main= "prism.drug.means")
abline(h=c(0,prism.drug.means.row.q), col=c("black", "darkblue", "lightblue", "orange", "red"))

prism.drug.means.row.var= apply(prism.drug.means, 1, function(x){var(x, na.rm= T)})
plot(prism.drug.means.row.var,  ylab= "row variance", xlab="treatment mean", main= "prism.drug.means")

prism.drug.means.kidney.row.mean= rowMeans(prism.drug.means.kidney, na.rm=T)
plot(prism.drug.means.kidney.row.mean,  ylab= "row means", xlab="treatment mean", main= "prism.drug.means.kidney")
abline(h=c(0,prism.drug.means.kidney.row.q), col=c("black", "darkblue", "lightblue", "orange", "red"))

prism.drug.means.kidney.row.var= apply(prism.drug.means.kidney, 1, function(x){var(x, na.rm= T)})
plot(prism.drug.means.kidney.row.var,  ylab= "row variance", xlab="treatment mean", main= "prism.drug.means.kidney")
#abline(h=prism.drug.means.kidney.row.q, col=c("darkblue", "lightblue", "orange", "red"))


#Histo kidney cancer cl mean wirkung in qX
hist(prism.drug.means.row.mean, breaks=50);
  abline(v= prism.drug.means.row.q, col=c("darkblue", "lightblue", "orange", "red"))
```







DATA CLEANUP
```{r}
rownames(prism.cl)= prism.cl$DepMap_ID
prism.cl=prism.cl[-c(1)]
prism.cl=prism.cl[,c(1,12,11,15,16,5,19,6,18)]

```



















```{r}
#rownumbers of prism.cl with Kidney Cancer
which(prism.cl$disease== "Kidney Cancer")

cl.kidney= prism.cl[c(which(prism.cl$disease== "Kidney Cancer")),]
```


```{r}
#kidney cancer exclusive df from prism.drug.means
prism.drug.means.kidney= prism.drug.means[rownames(cl.kidney),]

prism.drug.means.kidney.col.mean= apply(prism.drug.means.kidney, 2, function(x){mean(x, na.rm= T)})
hist(prism.drug.means.kidney.col.mean, breaks= 100);
abline(v=prism.drug.means.kidney.q, col=c("darkblue", "lightblue", "orange", "red"))

prism.drug.means.kidney.col.var= apply(prism.drug.means.kidney, 2, function(x){sd(x, na.rm= T)})
hist(prism.drug.means.kidney.col.var, breaks= 100)

pheatmap(prism.drug.means.kidney, show_colnames= F, show_rownames= F, main= " cell line viability of prism.drug.means.kidney")

```


```{r}
idx.kidney= colnames(prism.drug.means.kidney[which(apply(prism.drug.means.kidney, 2, min, na.rm=T)< prism.drug.means.kidney.q[1])])
unique(prism.treat[prism.treat$broad_id %in% idx.kidney,]$moa)
```



```{r}
prism.drug.means.kidney.q2.indication= which(table(prism.treat[prism.treat$broad_id %in% idx.kidney,]$disease.area)!=0)
dotchart(prism.drug.means.kidney.q2.indication, main= "indications of drugs in
         prism.drug.means.kidney in q2")
```





```{r}
prism.drug.means.kidney.q2.disease= which(table(prism.treat[prism.treat$broad_id %in% idx.kidney,]$disease.area)!=0)
dotchart(prism.drug.means.kidney.q2.indication, main= "indicitatons of drugs in
         prism.drug.means.kidney in q2")
```







```{r}
prism.kidney= prism[rownames(cl.kidney),]
prism.kidney.q= quantile(prism.kidney, c(0.02, 0.05, 0.1, 0.25), na.rm=T )
idx.prism.kidney= colnames(prism.kidney[which(apply(prism.kidney, 2, min, na.rm=T)< prism.kidney.q[1])])
#unique(prism.treat[prism.treat$broad_id %in% idx.kidney,]$moa)

prism.kidney.q2.disease_area= which(table(prism.treat[rownames(prism.treat) %in% idx.prism.kidney,]$disease.area)!=0)
dotchart(prism.kidney.q2.disease_area, main= "Disease area of treatments in prism.kidney in q2", ylim=1)


prism.kidney.q2.phase= table(prism.treat[rownames(prism.treat) %in% idx.prism.kidney,]$phase)
dotchart(prism.kidney.q2.phase, main= "Phase of treatments in prism.kidney in q2")


```



EFFIZIENZ DER DRUG/TREATMENT KLASSEN
```{r}
prism.kidney.q2.disease_area0= table(prism.treat[rownames(prism.treat) %in% idx.prism.kidney,]$disease.area)
a= prism.kidney.q2.disease_area0
b= c(table(prism.treat$disease.area))
```

```{r}

c= ((a/b)*100)
d=c[-(which(c==0))]
e=d[1:28]
```

```{r}
# sort by size
sorted_e <- sort(e)

# changing sequence by size
sorted_names <- names(sorted_e)
sorted_names_reordered <- reorder(sorted_names, sorted_e)

dotchart(sorted_e, labels = sorted_names_reordered, main= "Efficacy of of treatments by disease area in kidney cells ")

```



NEU
inds gives us specific conc per drug (columns)
```{r}
#dosage indices
i.treat= c(1,8) # choose dosage
n.drugs= 10 #number of drugs
n.all.treat= 8 #n of different conc per drug
inds= c()
for (i in 0:(n.drugs-1)) {
  inds= c(inds, i.treat + n.all.treat*i)
}

```

```{r}
# prism.kidney10= prism.kidney[,inds] #prism.kidney with first 2 conc per drug, n.drug=10
```

```{r}
# get small portion of prism.treat
prism.treat10= prism.treat[inds,]
# get list of all targets in prism.treat10
prism.treat.target10= prism.treat10$target
target.list10= unique(unlist(strsplit(as.character(prism.treat.target10),", "), recursive= T))
target.list10= target.list10[!is.na(target.list10)]

# get lists of cell types for which data exists in prism.kidney and prism.achilles
prism.kidney.celltypes= rownames(cl.kidney)
prism.achilles.celltypes= rownames(prism.achilles)
# get mask for prism.kidney which indicates if data for this zelltype exists in prism.achilles
prism.kidney.allDataMask= prism.kidney.celltypes %in% prism.achilles.celltypes

# get small portion of prism and prism.achilles data:
#  - only treatments as specified in inds
#  - only cell types for kidney cancer
#  - only cell types for which knockoff- AND treatment-viability data exists
prism.kidney10= prism.kidney[prism.kidney.allDataMask, inds]
prism.achilles.kidney10= prism.achilles[prism.achilles.celltypes %in% prism.kidney.celltypes, target.list10]

# sort rows of prism.kidney10 and prism.achilles.kidney10 for better comparability
prism.kidney10= prism.kidney10[sort(rownames(prism.kidney10)),]
prism.achilles.kidney10= prism.achilles.kidney10[sort(rownames(prism.achilles.kidney10)),]
```

```{r}
# temp.targetGene= target.list10[1]  # choose target gene for evaluation
temp.targetGene= "NFKB2"
cat("selected target gene= ", temp.targetGene, "\n\n")
temp.viab.KO10= prism.achilles.kidney10[temp.targetGene]  # get knockoff-viability data for target gene
temp.viabRel.treat10= prism.kidney10  # initialize
for (i in 1:(dim(prism.kidney10)[2])) {
  temp.viabRel.treat10[i] = temp.viabRel.treat10[i] / temp.viab.KO10
}
# reduce length of treatment names
temp.prism.kidney10= prism.kidney10
colnames(temp.prism.kidney10)= substring(colnames(temp.prism.kidney10), 5, nchar(colnames(temp.prism.kidney10))-8)
pheatmap(as.matrix(temp.prism.kidney10), cluster_rows=F, cluster_cols=F)  # heatmap of prism.kidney10 without knockoff reference
# reduce length of treatment names , todo: visually mark NA values
colnames(temp.viabRel.treat10)= substring(colnames(temp.viabRel.treat10), 5, nchar(colnames(temp.viabRel.treat10))-8)
pheatmap(as.matrix(temp.viabRel.treat10), cluster_rows=F, cluster_cols=F)  # todo: visibly mark NA values

# print heatmap of prism.achilles.kidney10 of which 1 column is used as divisor (reference)
pheatmap(as.matrix(prism.achilles.kidney10), cluster_rows=F, cluster_cols=F)

print(prism.treat10$target)

```

```{r}
# prism.treat10= prism.treat[inds,]
# prism.treat.target10= prism.treat10$target
# target.list10= unique(unlist(strsplit(as.character(prism.treat.target10),", "), recursive= T))
# target.list10= target.list10[!is.na(target.list10)]
```



list of all target genes of prism.kidney10
```{r}
prism.achilles.kidney10= prism.achilles.kidney[,target.list10]
```











