---
title: "R Notebook"
output: html_notebook
---
# data cleanup prism.cl
```{r}
#reorder prism.cl
colnames(prism.cl)
prism.cl<-prism.cl[,c(1,2,13,12,16,17,6,20,7,19)]

##make cell lines as row names
rownames(prism.cl)=prism.cl$DepMap_ID
prism.cl=prism.cl[-c(1)]
```

# special subdata extraction

## find kidney cancer cell lines in datasets

```{r kidney_datasets}

#rownumbers of prism.cl_cleaned with Kidney Cancer
which(prism.cl$disease== "Kidney Cancer")

prism.cl.kidney= prism.cl[c(which(prism.cl$disease== "Kidney Cancer")),]

#Gives back all cell lines which are kidney-cancer-cell lines
kidney_celllines<-print(prism.cl.kidney[,1])

#Gives back all cell line ACh-numbers which are cancer-cell lines
kidney_celllines_ACH <-print(rownames(prism.cl.kidney))

#stores only kidney cancer cell lines from other datasets in a new one
prism.exp.kidney <- prism.exp[rownames(prism.exp) %in% prism.cl.kidney[,1], ] 

#for prism dataset
prism.kidney <- prism[rownames(prism) %in% kidney_celllines_ACH,]

#for prism.achilles
prism.achilles.kindney<- prism.achilles[rownames(prism.achilles) %in% kidney_celllines_ACH,]
```

## Cancer type vectors

```{r}
#Vektoren mit einzelnen cancertypes erstellen
unique(prism.cl$disease)
cancertypes<- c(unique(prism.cl$disease))
```

# New cancer type specific datasets of prism.exp

```{r}
# Create a list to store data frames for each cancer type
prism_exp_list <- list()

# Loop over the cancertypes vector
for (cancer_type in cancertypes) {
  # Subset the prism.cl data frame for the current cancer type
  cancer_df <- prism.cl[which(prism.cl$disease == cancer_type), ]
  
  # Subset the prism.exp data frame based on the row names from the cancer_df
  prism_exp_subset <- prism.exp[rownames(cancer_df), ]
  
  # Store the subsetted prism.exp in the list with the cancer type as the list name
  prism_exp_list[[cancer_type]] <- prism_exp_subset
}

# Access the data frames for specific cancer types from the list
prism_exp_list[["Sarcoma"]]
# Access other cancer types similarly
```

#prism nach cancertypes einzelne Dataframes und topvarianten (q75) Gene der jeweiligen Cancertypes

```{r}
prism_NA_0 <- prism
prism_NA_0[is.na(prism_NA_0)] <- 0 # ersetze alle NA durch 0
# Create a list to store data frames for each cancer type
prism_list <- list()

# Loop over the cancertypes vector
for (cancer_type in cancertypes) {
  # Subset the prism.cl data frame for the current cancer type
  cancer_df <- prism.cl[which(prism.cl$disease == cancer_type), ]
  
  # Subset the prism.exp data frame based on the row names from the cancer_df
  prism_subset <- prism_NA_0[rownames(cancer_df), ]
  
  # Store the subsetted prism.exp in the list with the cancer type as the list name
  prism_list[[cancer_type]] <- prism_subset
}

# Access the data frames for specific cancer types from the list
prism_list[["Sarcoma"]]
# Access other cancer types similarly


# Loop over the cancertypes vector
for (cancer_type in names(prism_list)) {
  
  # Subset the prism.exp data frame based on the row names from the cancer_df
  prism.var<- apply(prism,MARGIN=2,FUN=var, na.rm=TRUE)
  prism.topvar=prism[,which(prism.var>=quantile(prism.var,probs=0.75,na.rm=TRUE))]
  
  prism.topvar.NA_0<-prism.topvar
  prism.topvar.NA_0[is.na(prism.topvar.NA_0)]<-0
  # Store the subsetted prism.exp in the list with the cancer type as the list name
  prism.var_list[[cancer_type]] <- prism.topvar.NA_0
}
#retrieve 75% var-quantile for each cancer type:
prism.var_list[["Sarcoma"]]

```

#Korrelation der hoch varianten Gene der spezifischen Cancertypes mit der copynumbervariation
#silhouette score für jeden der Varianz-Vektoren
```{r}
silhouette_list<-list()
for(cancer_type in names(prism.var_list)){
  distance<-dist(t(prism.topvar.NA_0))
  library(cluster)
  km_var<-kmeans(x=t(prism.topvar.NA_0), centers = 530, nstart = 10)
  s_var=silhouette(km_var$cluster,distance)
  silhouette_list[[cancer_type]]<-s_var
}
plot(silhouette_list[["Sarcoma"]])
#k=530-> s=0.06
#k=400-> s=0.05
#k=600, n 30-> s=0.06

silhouette_list<-list()
for (cancer_type in names(prism.var_list)){
  distance <- dist(t(prism.topvar.NA_0))
  library(cluster)

  for (n_centers in c(1,20,70,150,300,500,700,1000)) {
    km_var<-kmeans(x=t(prism.topvar.NA_0), centers = n_centers, nstart = 5)
    s_var=silhouette(km_var$cluster,distance)
    silhouette_list[[cancer_type]]<-s_var
}}
list(n_centers,silhouette_list[["Sarcoma"]])
  

```

#selbes Prozedere mit Elbow method
```{r}
#run k-means for k=5,50,100,300,500,700,1000 clusters, and for each k check the WSS value. How does WSS evolve with increasing k?


wss_list<-list()
#calculate within-square-distance
for(cancer_type in names(prism.var_list)){
  wss = sapply(c(5,50,100,300,500,700,1000),function(k) { 
    kmeans(x=t(prism.topvar.NA_0), centers =k)$tot.withinss
})
  wss_list[[cancer_type]]<-wss
}
plot(c(5,50,100,300,500,700,1000),wss_list[["Kidney Cancer"]],type='b',pch=19,xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")

##für welchen cancer type ist dieser plot jetzt? ->for-Schleife schreiben und Liste mit den einzelnen wss-zu-k-Werten für die einzelnen cancertypes kreieren
```

#für kidney cancer
```{r}
prism.var_list[["Kidney Cancer"]]
#->ca. 100 center beim clustering für 2782 Gene-> wonach wurde geclustert? wie kann ich mir den Inhalt eines Clusters anzeigen lassen?

km_kidney_cluster = kmeans(t(prism.var_list[["Kidney Cancer"]]), 
            centers = 100, 
            nstart = 10)
print(km_kidney_cluster)


subset(prism.var_list[["Kidney Cancer"]], km_kidney_cluster$cluster == 30) 

kmeans_fit_Kidney <- kmeans(prism.var_list[["Kidney Cancer"]], centers = 100) # Anzahl der Cluster anpassen
head(kmeans_fit_Kidney$cluster)

#datframe mit wirkstoffnamen pro cluster
dim(subset(prism.var_list[["Kidney Cancer"]], kmeans_fit_Kidney$cluster == 70)) #für Cluster 70.

#will ich, dass nach Zellinien oder nach Wirkstoffen geclustert wird?--> nach Wirkstoff, da wir ja nur in den Kidney cell lines suchen
#warum gibt der mir so viele NA-Werte?
#wie kann ich mir jetzt die entsprechenden mode of actions ausgeben lassen?-> drugID (colname im prism.var_list dataframe) in prism.treat suchen und prism.treat$moa ausgeben lassen
```




#wie kann ich eine matrix aus der Vektor-Liste mit zugehörigem cancertype machen?
```

# Data lookup

## prism.exp

This dataset shows the Transcripts per million of the genes (columns) for the cell lines (rows).

```{r}
head(prism.exp)

dim(prism.exp)

class(prism.exp)
```

```{r}
#compute genes with highest variance in expression
prism.exp.var<- apply(prism.exp,MARGIN=2,FUN=var)
```

#### Range of the values

##### Code

```{r}
range(prism.exp)

range(prism.exp.kidney)
```

##### Plot

```{r, fig.width=8, fig.height=6}

hist(unlist(prism.exp), breaks = 23, main = "Histogram of prism.exp", xlab = "TPM value")
hist(unlist(prism.exp.kidney), breaks = 23, main = "Histogram of prism.exp.kidney", xlab = "TPM value")

density(unlist(prism.exp))
density(unlist(prism.exp.kidney))

#sum(unlist(prism.exp) >= 0 & unlist(prism.exp) <= 1)

```

```{r}
pheatmap(prism.exp.kidney,show_colnames = FALSE, main = "Heatmap of prism.exp.kidney: TPM")
```

```{r}
#Plot for the first i genes the frequency of expression values
for (i in 1:10) {
  hist(prism.exp[, i], main = paste0("Histogram for column ", i), breaks = 200,)
}
```

#### Summary for cell lines

##### Code

```{r}
rm(prism.exp_row_summaries)

# create a function to calculate the mean, median, and quantiles for a row/cell lines
row_summary <- function(row) {
  result <- c(mean(row), median(row), quantile(row, c(0.25, 0.75)), min(row), max(row), IQR(row))
  names(result) <- c("mean", "median", "q1", "q3", "min", "max", "IQR")
  return(result)
}

# apply the function to each row using apply()
prism.exp.row_summaries <- apply(prism.exp, 1, row_summary)

# convert the result to a data frame
prism.exp.row_summaries_df <- as.data.frame(t(prism.exp.row_summaries))

hist(prism.exp.row_summaries_df$q1, breaks = 50, xlab =  "Q25", main = "Histogram for 25%-Q of cell lines")
hist(prism.exp.row_summaries_df$q3, breaks = 80, xlab = "Q75", main = "Histogram for 75%-Q of cell lines")
hist(prism.exp.row_summaries_df$IQR, breaks = 80, xlab="IQR", main = "Histogram for IQR of cell lines")

plot(prism.exp.row_summaries_df$IQR, ylab = "IQR" , xlab =  "cell lines", main = "Values of IQR of cell lines")
```

##### Plot

```{r}
par(plt=c(0,0,1,1)+1,  # Set plot region to the whole device
    mar=c(2,3,4,1) + 0.1, # Set margin sizes
    xpd=TRUE) # Allow plotting outside the plot region
plot(prism.exp.row_summaries_df$q1, 
     ylab = "expression values of quantiles",
     xlab = "cell lines", 
     main = "Values of 25%-Q and 75%-Q of cell lines", 
     col = "red4",
     cex = 1.5,
     ylim = c(min(prism.exp.row_summaries_df$q1, prism.exp.row_summaries_df$q3), 
              max(prism.exp.row_summaries_df$q1, prism.exp.row_summaries_df$q3)))
points(prism.exp.row_summaries_df$q3, col="navy") #im gleichen Diagramm auch 75%
```

```{r, PCA_for_prism.exp}

pca = prcomp(t(prism.exp), center = FALSE, scale = FALSE)
pca$sdev
print(pca$sdev[0:66])
colnames(prism.exp.var %% pca$sdev[0:66])
```

## prism.treat

The dataset shows for each treatment (rows) information about the drugs and dosis.

```{r summary_prism.treat}
summary(prism.treat)
```

## prism

This dataset shows the treatment effects (columns) on the cell growth for each cell line (rows).

```{r}
head(prism)
```

```{r}
hist(unlist(prism), main= "Histogram of prism", xlab="Effect of treatment on cell growth ", breaks= 50)
```

```{r}
plot(prism[,100], xlab = "cell lines", ylab= "treatment effect", main= "Effect of one treatment on all cell lines")
abline(h = 0, col="red")
```

```{r}
#NA Werte looki looki
colnames(prism)[1]
sum(is.na(prism$"BRD-A00077618-236-07-6::0.00061034::HTS002"))
```

```{r}
 #Berechne/plot den Mittelwert/SD jeder Spalte des Dataframes, wobei NA-Werte ignoriert werden

prism.col.mean= apply(prism, 2, function(x){mean(x, na.rm= T)})
plot(prism.col.mean, ylab= "column means", xlab="treatments", main= "prism" )

prism.col.var= apply(prism, 2, function(x){var(x, na.rm= T)})
plot(prism.col.var, ylab= "column variance", xlab="treatments", main= "prism" )
```

```{r}
#Funktion, welche  columns nach parameter drug extrahiert in noch leeres df
extract_columns <- function(prism, drug) {
  col_indices <- grep(drug, colnames(prism))
  
# Erstelle ein neues Dataframe mit ausgewählten Spalten
new_prism <- prism[, col_indices]
  
return(new_prism)
}
```

```{r}
#extrahiert drug name // hätte man auch mit prism.treat machen können
drugs= unique(sapply(strsplit(colnames(prism),"::"),`[`, 1))


```

## prism.achilles

This dataset shows gene knockout-scores, representing the importance of one gene for cell survival, of every gene (columns) for the cell lines (rows).

```{r}
head(prism.achilles)
```

```{r}
range(prism.achilles, na.rm=TRUE)

hist(unlist(prism.achilles), breaks = 80, main = "Histogram of prism.achilles", xlab = "gene knock-out scores")
```

## prism.cnv

```{r}
range(prism.cnv)
```

```{r}
hist(unlist(prism.cnv), main= "Histogram of prism.cnv", xlab="copy number variation", breaks=300)
```

# Correlation between cnv & expression
```{r}
#rownames(prism.exp %in% prism.cnv)
#This is because the rownames() function is used to retrieve the row names of a data frame or matrix, but the %in% operator returns a logical vector indicating which elements of the left-hand side vector are also in the right-hand side vector. The output of %in% is not a data frame or matrix, but a logical vector that does not have row names. Therefore, when you apply rownames() to the result of the %in% operation, it returns NULL indicating that there are no row names associated with the logical vector.
#prism.exp %in% prism.cnv

#identify missing rows in exp that are present in cnv
diff_rows <- setdiff(rownames(prism.cnv), rownames(prism.exp))
diff_rows <- unique(c(diff_rows, setdiff(rownames(prism.exp), rownames(prism.cnv))))
print(diff_rows)

#prism.cnv_corr_cleaned<-prism.cnv[-c("ACH-000309","ACH-000047","ACH-000979","ACH-001024")]
#which(rownames(prism.cnv)=="ACH-001024")
#85, 268, 294,307
prism.cnv_corr_cleaned<- prism.cnv[-c(85,268,294,307),]

#identify missing columns that are present in cnv but not in exp
diff_cols <- setdiff(colnames(prism.cnv), colnames(prism.exp))
diff_cols <- unique(c(diff_cols, setdiff(colnames(prism.exp), colnames(prism.cnv))))
print(diff_cols)

diff_cols2 <- setdiff(colnames(prism.exp), colnames(prism.cnv))
diff_cols2<- unique(c(diff_cols, setdiff(colnames(prism.cnv), colnames(prism.exp))))
prism.cnv_corr_cleaned<-prism.cnv_corr_cleaned[-which(colnames(prism.cnv) %in% diff_cols)]
prism.exp_corr_cleaned<-prism.exp[-which(colnames(prism.exp) %in% diff_cols2)]

#reorder rows in an ascending way
#1.step:rownames(prism.cnv_corr_cleaned)
prism.cnv_corr_cleaned<- prism.cnv_corr_cleaned[order(rownames(prism.cnv_corr_cleaned)),]
prism.exp_corr_cleaned<- prism.exp_corr_cleaned[order(rownames(prism.exp_corr_cleaned)),]

#reorder columns by names
prism.cnv_corr_cleaned= prism.cnv_corr_cleaned[,colnames(prism.exp_corr_cleaned)]

#colnames(prism.cnv_corr_cleaned)

exp_cnv_corr<-cor(prism.exp_corr_cleaned, prism.cnv_corr_cleaned, method= c("spearman"))
sum(exp_cnv_corr> 0.5)
sum(exp_cnv_corr< "-0.5")
#keine Korrelation +0.5 zwischen Expression und copy number variation
#512969 < -0.5

#correlation nur für hohe cnv berechnen:
highcnv<- which(prism.cnv_corr_cleaned> 2, arr.ind=TRUE)

exp_cnv_corr[highcnv]
sum(exp_cnv_corr[highcnv]>0.5)
sum(exp_cnv_corr[highcnv]< "-0.5")
#11 Werte >0.5, 9962 Werte/17906 < -0.5
#starke negative korrelation für hohe cnv mit exp?


##for loop to only correlate the gene with itself and not with the other genes
# for x in data frame1 correlate with x in data frame 2
```

# clustering of similar expression profiles in cl
```{r}
#dataset austauschen
#compute genes with highest variance in expression
prism.exp.var<- apply(prism.exp,MARGIN=2,FUN=var)

#top25% with highest variance:
q75.exp.var = quantile(prism.exp.var, probs = 0.75)   # what is the 75% quantile of the variance?
i.topvar.exp = which(prism.exp.var >= q75.exp.var)   # all columns (genes) with variance higher than 75%

prism.exp.topvar = prism.exp[,i.topvar.exp]   # only select the gene with a variance in the top 25%

print(prism.exp.topvar)

#k-means clustering
km = kmeans(x=t(prism.exp.topvar), 
            centers = 23, 
            nstart = 10)
#Centers=k; nstart= how many random sets should be chosen?

#run k-means for k=1 to k=40 clusters, and for each k check the WSS value. How does WSS evolve with increasing k?
##Elbow method
#calculate within-square-distance
wss = sapply(1:40,function(k) { 
  kmeans(x=t(prism.exp.topvar), centers =k)$tot.withinss
})
plot(1:40,wss,type='b',pch=19,xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")


#calculate for each tumor
#calculate variance fpor each of the tumors
##highly variable gene
#run silhouette score

D = dist(t(prism.exp.topvar))
library(cluster)
km1 = kmeans(x=t(prism.exp.topvar), centers = 2, nstart = 10)
s = silhouette(km1$cluster,D)
plot(s)

#3 cluster: si=0.21, 23 cluster: si=0.04, 4:0.16, 2:0.33, 
#average silhouette width für mehrere k s plotten und knick in kurve suchen
```
#NA-->0
prism.kidney_NA_0<-prism.kidney
prism.kidney_NA_0[is.na(prism.kidney_NA_0)] <- 0

# Step 1: Calculate variances
prism.kidney.cellline_var <- apply(prism.kidney_NA_0, 1, var)

# Step 2: Set threshold for upper n percent
n_percent <- 0.1
threshold <- quantile(prism.kidney.cellline_var, 1 - n_percent)

# Step 3: Filter highly variant pairs
highly_variant_celllines <- rownames(prism.kidney_NA_0)[which(prism.kidney.cellline_var > threshold)]

print(highly_variant_celllines)
```

### of drugs- kidney cancer

```{r}
# Step 1: Calculate variances
prism.kidney.drugs_var <- apply(prism.kidney_NA_0, 2, var)

# Step 2: Set threshold for upper n percent
n_percent <- 0.0001
threshold <- quantile(prism.kidney.drugs_var, 1 - n_percent)

# Step 3: Filter highly variant pairs
highly_variant_drugs <- colnames(prism.kidney_NA_0)[which(prism.kidney.drugs_var > threshold)]

print(highly_variant_drugs)
```

#### find them in prism.treat

```{r}

# Identify the highly variant drug names
highly_variant_drugs_kidney <- colnames(prism.kidney_NA_0)[which(prism.kidney.drugs_var > threshold)]

# Show the filtered rows
print(prism.treat[row.names(prism.treat) %in% highly_variant_drugs_kidney, ])

rm(prism.kidney.cellline_variances)
```

#### extract dataset of prism for those cell lines

```{r}
#store the drugnames as vector
highly_variant_drugs_kidney_drugnames<- rownames(prism.treat[row.names(prism.treat) %in% highly_variant_drugs_kidney, ])

# Extract the subset of columns based on the highly variant drug names
prism.kidney.highly_variant_drugs <- prism[, colnames(prism) %in% highly_variant_drugs_kidney_drugnames]

# Print the new dataset
print(prism.kidney.highly_variant_drugs)

```

#### show the cell lines with the most extreme values

```{r}
#NA=0
prism.kidney.highly_variant_drugs.NA_0<-prism.kidney.highly_variant_drugs
prism.kidney.highly_variant_drugs.NA_0[is.na(prism.kidney.highly_variant_drugs.NA_0)] <- 0  

# Find the upper n% extreme values 


# Extract the cell line names as vector


```

### of drugs - all cell lines

```{r}
#NA-->0
prism.NA_0<-prism
prism.NA_0[is.na(prism.NA_0)] <- 0

# Step 1: Calculate pair variances
prism.drugs_var <- apply(prism.NA_0, 2, var)

# Step 2: Determine threshold for upper n percent
n_percent <- 0.0005
threshold <- quantile(prism.drugs_var, 1 - n_percent)

# Step 3: Filter highly variant pairs
highly_variant_drugs <- colnames(prism.NA_0)[which(prism.drugs_var > threshold)]

print(highly_variant_drugs)

# Identify the highly variant drug names
highly_variant_drugs <- colnames(prism.NA_0)[which(prism.drugs_var > threshold)]

# Show the filtered rows
print(prism.treat[row.names(prism.treat) %in% highly_variant_drugs, ])
```

## highest prism values

### for kidney cancer

```{r}
#Range of prism-values
range(prism.kidney_NA_0)

#Set threshold
prism_threshold_kidney<- -9
                                                
# Create an empty vector to store the column names
high_effect_drugs_kidney <- c()

# Iterate over each column of the dataset
for (col in colnames(prism.kidney_NA_0)) {
  # Check if any value in the column is lower than the threshold
  if (any(prism.kidney_NA_0[, col] < prism_threshold_kidney)) {
    # If true, append the column name to the vector
    high_effect_drugs_kidney <- c(high_effect_drugs_kidney, col)
  }
}

#Show the filtered rows
print(prism.treat[rownames(prism.treat) %in% high_effect_drugs_kidney,])
```

### for all cell lines

```{r}
#Range of prism-values
range(prism.NA_0)

#Set threshold
prism_threshold<- -11
                                                
# Create an empty vector to store the column names
high_effect_drugs <- c()

# Iterate over each column of the dataset
for (col in colnames(prism.NA_0)) {
  # Check if any value in the column is lower than the threshold
  if (any(prism.NA_0[, col] < prism_threshold)) {
    # If true, append the column name to the vector
    high_effect_drugs <- c(high_effect_drugs, col)
  }
}

#Show the filtered rows
print(prism.treat[rownames(prism.treat) %in% high_effect_drugs,])
```

# which cell lines are mostly effected by those drugs

### for kidney cancer

# copy number variation and cancer types

```{r}
# Calculate the average copy number per gene
average_copy_number <- rowMeans(prism.cnv)

# Set the threshold for high amplification
threshold <- 1.01  # Adjust this threshold as desired

# Identify genes that are highly amplified
highly_amplified_genes <- names(average_copy_number[average_copy_number > threshold])

# Print the highly amplified genes
print(highly_amplified_genes)

```

**fix this! -doesn't work as expected:**

```{r}
# Create a data frame to store the results
amplified_genes_by_cancer <- data.frame()

# Iterate over each unique cancer type in prism.cl
for (disease in unique(prism.cl$disease)) {
  # Subset the prism.cl data frame for the current cancer type
  cancer_subset <- prism.cl[prism.cl$disease %in% disease, ]
  
  # Get the cell lines associated with the current cancer type
  cell_lines <- cancer_subset$cell_line
  
  # Subset the prism.cnv data frame for the cell lines of the current cancer type
  cnv_subset <- prism.cnv[]
  
  # Calculate the average copy number per gene for the current cancer type
  average_copy_number <- rowMeans(cnv_subset)
  
  # Define a threshold value for gene amplification for the current cancer type
  amplification_threshold <- 1.0  # Adjust this threshold as needed
  
  # Identify genes that are highly amplified for the current cancer type
  highly_amplified_genes <- names(average_copy_number[average_copy_number > amplification_threshold])
  
  # Create a data frame to store the results for the current cancer type
 result <- data.frame(CancerType = disease, AmplifiedGenes = paste(highly_amplified_genes, collapse = ", "))

  
  # Append the result to the amplified_genes_by_cancer data frame
  amplified_genes_by_cancer <- rbind(amplified_genes_by_cancer, result)
}

# Print the results
print(amplified_genes_by_cancer)

```
