---
title: "topic05_team02_report"
author: "Jessica Albrecht, Franca Klaeren & Arne Weigel"
date: "2023-07-05"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: 4
---

# Report

## Introduction

Beschreibung der Fragestellung, Vorstellung der publizierten Vorarbeiten,\...

## Material and Methods

```{r load_datasets, include=FALSE}
load("./cellline_datasets.rda")
load("./prism_datasets.rda")
```

The data we used for the project was published by XXXX.

bitte hier die Referenzen der Publikationen/Webseiten, aus denen die Daten stammen, und die Beschreibung der Methoden, die benutzt wurden.

### data cleanup

Before we started to work with our data sets, we first had to do some data cleanup. For that we generated some sub-data sets for the 17 kidney cancer cell lines and for some areas of application we needed to set NA-values to zero.

```{r cleanup_cols_prism.cl, include=FALSE}
#clean up of columns of prism.cl
rownames(prism.cl)= prism.cl$DepMap_ID
prism.cl=prism.cl[-c(1)]
prism.cl=prism.cl[,c(1,12,11,15,16,5,19,6,18)]
```

```{r kidney_cancer_subsets, include=FALSE}
#rownumbers of prism.cl with Kidney Cancer
which(prism.cl$disease== "Kidney Cancer")

prism.cl.kidney= prism.cl[c(which(prism.cl$disease== "Kidney Cancer")),]

#Gives back all cell lines which are kidney-cancer-cell lines
kidney_celllines<-print(prism.cl.kidney[,1])

#Gives back all cell line ACh-numbers which are cancer-cell lines
kidney_celllines_ACH <-rownames(prism.cl.kidney)

#stores only kidney cancer cell lines from other datasets in a new one
prism.exp.kidney <- prism.exp[kidney_celllines_ACH, ]

#for prism dataset
prism.kidney <- prism[rownames(prism) %in% kidney_celllines_ACH,]

#for prism.achilles
prism.achilles.kidney<- prism.achilles[rownames(prism.achilles) %in% kidney_celllines_ACH,]
```

```{r NA_0, include=FALSE}
#set NA to 0
prism_NA_0 <- prism
prism_NA_0[is.na(prism_NA_0)] <- 0 

prism.kidney_NA_0<-prism.kidney
prism.kidney_NA_0[is.na(prism.kidney_NA_0)] <- 0


prism.achilles.kidney_NA_0<-prism.achilles.kidney
prism.achilles.kidney_NA_0[is.na(prism.achilles.kidney_NA_0)] <- 0
```

## Results

Darstellung der Ergebnisse und Abbildungen

+------------------------------------------------------------+----------------+
| Must-haves                                                 |                |
+============================================================+================+
| descriptive statistics including graphical representations | mean           |
|                                                            |                |
|                                                            | median         |
|                                                            |                |
|                                                            | **quantiles**  |
|                                                            |                |
|                                                            | **Histograms** |
|                                                            |                |
|                                                            | range          |
+------------------------------------------------------------+----------------+
| dimension reduction analysis                               | PCA            |
+------------------------------------------------------------+----------------+
| statistical tests                                          |                |
+------------------------------------------------------------+----------------+
| linear regression                                          |                |
+------------------------------------------------------------+----------------+

### descriptive statistics

For a quick overview of the content we printed the headlines of the data sets for example for prism.cl.

```{r head_prism.cl, echo=FALSE}
head(prism.cl)
```

We looked for the data distribution for prism.exp, prism.achilles,prism.cnv and prism which seemed to be not normal distributed. This conclusion was confirmed by a Anderson-Darling normality test (p-value \< 2.2e-16).

```{r distribution_histograms, echo=FALSE}
par(mfrow=c(2,2))
hist(unlist(prism), breaks = 60, main = "Histogram of prism", xlab = "treatment effect values")

hist(unlist(prism.exp), breaks = 60, main = "Histogram of prism.exp", xlab = "transcripts per million")

hist(unlist(prism.achilles), breaks = 60,main = "Histogram of prism.achilles", xlab = "gene knockout-scores")

hist(unlist(prism.cnv), main= "Histogram of prism.cnv", xlab="copy number variation", breaks=60)
```

```{r check_normal_distribution, include=FALSE}
#values as vectors
prism_vector <- unlist(prism)
prism.achilles_vector<-unlist(prism.achilles)
prism.exp_vector<- unlist(prism.exp)
prism.cnv_vector<-unlist(prism.cnv)

# Perform the Anderson-Darling normality test
library(nortest)
hyp_test_prism_distribution<- ad.test(prism_vector)
hyp_test_prism_distribution

hyp_test_prism.achilles_distribution<-ad.test(prism.achilles_vector)
hyp_test_prism.achilles_distribution

hyp_test_prism.exp_vector<-ad.test(prism.exp_vector)
hyp_test_prism.exp_vector

hyp_test_prism.cnv_vector<-ad.test(prism.cnv_vector)
hyp_test_prism.cnv_vector
```

The boxplot of the data shows small inter quartile ranges for prism.cnv and prism.achilles and a bigger range for the prism data set. This means we have more spreaded values for prism than for the other two data sets.

```{r boxplot_prism_cnv_ach, echo=FALSE}
boxplot(
  prism_vector,prism.cnv_vector, prism.achilles_vector,
  outline = FALSE,
  names=c("prism", "prism.cnv", "prism.achilles"))
```

### Do some high effective drugs target essential genes of kidney cancer?

#### Which genes are essential for kidney cancer cell lines?

```{r essential_genes, include=FALSE}
#old~new
#cnv~exp
#exp~ach
#identify missing rows in exp that are present in achilles
diff_rows <- setdiff(rownames(prism.exp.kidney), rownames(prism.achilles.kidney))
diff_rows <- unique(c(diff_rows, setdiff(rownames(prism.achilles.kidney), rownames(prism.exp.kidney))))
#print(diff_rows)

#prism.exp.kidney_corr_cleaned<-prism.exp.kidney
#"ACH-000046" "ACH-000171" "ACH-000313" "ACH-000428" "ACH-000429" "ACH-000457"
#which(rownames(prism.exp.kidney)=="ACH-000457")
#2,4,7,9,10,12
prism.exp.kidney_corr_cleaned<- prism.exp.kidney[-c(2,4,7,9,10,12),]

#identify missing columns that are present in prism.exp.kidney_corr_cleaned but not in achilles
diff_cols <- setdiff(colnames(prism.exp.kidney), colnames(prism.achilles.kidney))
diff_cols <- unique(c(diff_cols, setdiff(colnames(prism.achilles.kidney), colnames(prism.exp.kidney))))
#print(diff_cols)


diff_cols2 <- setdiff(colnames(prism.achilles.kidney), colnames(prism.exp.kidney))
diff_cols2<- unique(c(diff_cols, setdiff(colnames(prism.exp.kidney), colnames(prism.achilles.kidney))))
prism.exp.kidney_corr_cleaned<-prism.exp.kidney_corr_cleaned[-which(colnames(prism.exp.kidney) %in% diff_cols)]
prism.achilles.kidney_corr_cleaned<-prism.achilles.kidney[-which(colnames(prism.achilles.kidney) %in% diff_cols2)]

#reorder rows in an ascending way
#1.step:rownames(prism.exp.kidney_corr_cleaned)
prism.exp.kidney_corr_cleaned<- prism.exp.kidney_corr_cleaned[order(rownames(prism.exp.kidney_corr_cleaned)),]
prism.achilles.kidney_corr_cleaned<- prism.achilles.kidney_corr_cleaned[order(rownames(prism.achilles.kidney_corr_cleaned)),]

#reorder columns by names
prism.exp.kidney_corr_cleaned= prism.exp.kidney_corr_cleaned[,colnames(prism.achilles.kidney_corr_cleaned)]

correlation_ach_exp_kidney<-matrix()
for (g in colnames(prism.achilles.kidney_corr_cleaned)) {  
  valse_g_ach_exp_kidney = prism.achilles.kidney_corr_cleaned[,colnames(prism.achilles.kidney_corr_cleaned) == g]  
  valsc_g_ach_exp_kidney = prism.exp.kidney_corr_cleaned[,colnames(prism.exp.kidney_corr_cleaned) == g]  
  correlation_ach_exp_kidney[[g]]=cor(valse_g_ach_exp_kidney, valsc_g_ach_exp_kidney, method= c("spearman"))  
}  
final_correlation_ach_exp_kidney = t(data.frame(correlation_ach_exp_kidney))

colnames_pos_corr <- colnames(final_correlation_ach_exp_kidney)[which(final_correlation_ach_exp_kidney >= quantile(final_correlation_ach_exp_kidney, 0.98, na.rm = TRUE))]

colnames_neg_corr <- colnames(final_correlation_ach_exp_kidney)[which(final_correlation_ach_exp_kidney <= quantile(final_correlation_ach_exp_kidney, 0.02, na.rm = TRUE))]

high_corr_genenames<-c(colnames_neg_corr, colnames_pos_corr)
```

```{r correlation_histogram, echo=FALSE}
hist(correlation_ach_exp_kidney, breaks = 40)
abline(v=c(-0.9,0.9))
```

#### Which drugs do have a high negative effect on the cell lines?

```{r neg_effect_drugs, include=FALSE}
#Set threshold depending on the distribution of prism 
prism.kidney_threshold<- quantile(prism.kidney, 0.02, na.rm=TRUE)

                                                
# Create an empty vector to store the column names
high_effect_drugs_kidney <- c()

# Iterate over each column of the dataset
for (col in colnames(prism.kidney_NA_0)) {
  # Check if any value in the column is lower than the threshold
  if (any(prism.kidney_NA_0[, col] <= prism.kidney_threshold)) {
    # If true, append the column name to the vector
    high_effect_drugs_kidney <- c(high_effect_drugs_kidney, col)
  }
}
#subset of prism.treat for drugs with inhibiting impact on kidney cancer cells
#Show/store the filtered rows
prism.treat.kidney.high_effect_drugs<-print(prism.treat[rownames(prism.treat) %in% high_effect_drugs_kidney,])


prism.treat.kidney.high_effect_drugs$target <- as.character(prism.treat.kidney.high_effect_drugs$target)

##Create a vector with all targets of the high_effect drugs

# Extract all targets as a single character vector
targetgenes_high_effect_drugs_kidney <- 
  unlist(strsplit(prism.treat.kidney.high_effect_drugs$target, ", "))

# Remove any leading/trailing white spaces and NA values
targetgenes_high_effect_drugs_kidneys <- trimws(targetgenes_high_effect_drugs_kidney)
targetgenes_high_effect_drugs_kidney <- targetgenes_high_effect_drugs_kidney[!is.na(targetgenes_high_effect_drugs_kidney)]

# Print the resulting vector of targets
#print(targetgenes_high_effect_drugs_kidney)
```

```{r venn_genes, echo=FALSE}
# Create a list of the gene names in each vector
set1 <- unique(high_corr_genenames)
set2 <- unique(targetgenes_high_effect_drugs_kidney)

list_venn1<-list(set1, set2)

# Create the Venn diagram
library(VennDiagram)
venn1<-venn.diagram(
  list_venn1 ,
  category.names = c("essential genes(high corr. btw. prism.ach & prism.exp)", "target genes of high effective drugs "),
  fill = c("lightblue", "steelblue4"),
  main ="venn diagram",
  main.fontface= "bold",
  sub = "essential genes vs.target genes of high effective drugs ",
  col= "white",
  cat.pos= 9,
  filename = NULL
)
grid.newpage()
grid.draw(venn1)
```

These genes are essential genes as well as targets of high inhibiting drugs:

```{r intersect_genes, echo=FALSE}
#find the genes in the intersection
intersect_data <- intersect(set1, set2)
intersect_data
```

## Discussion

kurze Zusammenfassung der Ergebnisse, Diskussion über gewonnen Erkenntnisse, Einschränkungen in der Analyse

## Referenzen
