---
title: "topic05_team02_report"
author: "Jessica Albrecht, Franca Klaeren & Arne Weigel"
date: "2023-07-05"
output:
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: 4
---

# Report

## Introduction

Beschreibung der Fragestellung, Vorstellung der publizierten Vorarbeiten,\...

## Material and Methods

```{r load_datasets, eval=FALSE, include=FALSE}
load("/cellline_datasets.rda")
load("/prism_datasets.rda")
```

The data we used for the project was published by XXXX.

bitte hier die Referenzen der Publikationen/Webseiten, aus denen die Daten stammen, und die Beschreibung der Methoden, die benutzt wurden.

### data cleanup

Before we started to work with our data sets, we first had to do some data cleanup. For that we generated some sub-data sets for the 17 kidney cancer cell lines and for some areas of application we needed to set NA-values to zero.

```{r cleanup_cols_prism.cl, include=FALSE}
#clean up of columns of prism.cl
rownames(prism.cl)= prism.cl$DepMap_ID
prism.cl=prism.cl[-c(1)]
prism.cl<-prism.cl[,c(1,12,11,15,16,5,19,6,18)]

```

```{r kidney_cancer_subsets, include=FALSE}
#rownumbers of prism.cl with Kidney Cancer
which(prism.cl$disease== "Kidney Cancer")

prism.cl.kidney= prism.cl[c(which(prism.cl$disease== "Kidney Cancer")),]

#Gives back all cell lines which are kidney-cancer-cell lines
kidney_celllines<-print(prism.cl.kidney[,1])

#Gives back all cell line ACh-numbers which are cancer-cell lines
kidney_celllines_ACH <-rownames(prism.cl.kidney)

#stores only kidney cancer cell lines from other datasets in a new one
prism.exp.kidney <- prism.exp[kidney_celllines_ACH, ]

#for prism dataset
prism.kidney <- prism[rownames(prism) %in% kidney_celllines_ACH,]

#for prism.achilles
prism.achilles.kidney<- prism.achilles[rownames(prism.achilles) %in% kidney_celllines_ACH,]
```

```{r NA_0, include=FALSE}
#set NA to 0
prism_NA_0 <- prism
prism_NA_0[is.na(prism_NA_0)] <- 0 

prism.kidney_NA_0<-prism.kidney
prism.kidney_NA_0[is.na(prism.kidney_NA_0)] <- 0


prism.achilles.kidney_NA_0<-prism.achilles.kidney
prism.achilles.kidney_NA_0[is.na(prism.achilles.kidney_NA_0)] <- 0
```

#### Highly variant gene expressions

```{r include=FALSE}
#Vektoren mit einzelnen cancertypes erstellen
unique(prism.cl$disease)
cancertypes<- c(unique(prism.cl$disease))

# Create a list to store data frames for each cancer type
prism_list <- list()

# Loop over the cancertypes vector
for (cancer_type in cancertypes) {
  # Subset the prism.cl data frame for the current cancer type
  cancer_df <- prism.cl[which(prism.cl$disease == cancer_type), ]
  
  # Subset the prism.exp data frame based on the row names from the cancer_df
  prism_subset <- prism_NA_0[rownames(cancer_df), ]
  
  # Store the subsetted prism.exp in the list with the cancer type as the list name
  prism_list[[cancer_type]] <- prism_subset
}

# Access the data frames for specific cancer types from the list
prism_list[["Sarcoma"]]
# Access other cancer types similarly

prism.var_list<-list()
# Loop over the cancertypes vector
for (cancer_type in names(prism_list)) {
  
  prism.var<- apply(prism_NA_0,MARGIN=2,FUN=var)
  prism.topvar=prism_NA_0[rownames(cancer_df),which(prism.var>=quantile(prism.var,probs=0.75,na.rm=TRUE))]
  
  prism.topvar.NA_0<-prism.topvar
  prism.topvar.NA_0[is.na(prism.topvar.NA_0)]<-0
  # Store the subsetted prism.exp in the list with the cancer type as the list name
  prism.var_list[[cancer_type]] <- prism.topvar.NA_0
}
#retrieve 75% var-quantile for each cancer type:
prism.var_list[[cancer_type]]
```

### Reordering of cnv and exp for correlation analysis

```{r}
#identify missing rows in exp that are present in cnv
diff_rows <- setdiff(rownames(prism.cnv), rownames(prism.exp))
diff_rows <- unique(c(diff_rows, setdiff(rownames(prism.exp), rownames(prism.cnv))))

#prism.cnv_corr_cleaned<-prism.cnv[-c("ACH-000309","ACH-000047","ACH-000979","ACH-001024")]
#which(rownames(prism.cnv)=="ACH-001024")
prism.cnv_corr_cleaned<- prism.cnv[-c(85,268,294,307),]

#identify missing columns that are present in cnv but not in exp
diff_cols <- setdiff(colnames(prism.cnv), colnames(prism.exp))
diff_cols <- unique(c(diff_cols, setdiff(colnames(prism.exp), colnames(prism.cnv))))

diff_cols2 <- setdiff(colnames(prism.exp), colnames(prism.cnv))
diff_cols2<- unique(c(diff_cols, setdiff(colnames(prism.cnv), colnames(prism.exp))))
prism.cnv_corr_cleaned<-prism.cnv_corr_cleaned[-which(colnames(prism.cnv) %in% diff_cols)]
prism.exp_corr_cleaned<-prism.exp[-which(colnames(prism.exp) %in% diff_cols2)]

#reorder rows in an ascending way
#1.step:rownames(prism.cnv_corr_cleaned)
prism.cnv_corr_cleaned<- prism.cnv_corr_cleaned[order(rownames(prism.cnv_corr_cleaned)),]
prism.exp_corr_cleaned<- prism.exp_corr_cleaned[order(rownames(prism.exp_corr_cleaned)),]

#reorder columns by names
prism.cnv_corr_cleaned= prism.cnv_corr_cleaned[,colnames(prism.exp_corr_cleaned)]
```

## Results

Darstellung der Ergebnisse und Abbildungen

+------------------------------------------------------------+----------------+
| Must-haves                                                 |                |
+============================================================+================+
| descriptive statistics including graphical representations | mean           |
|                                                            |                |
|                                                            | median         |
|                                                            |                |
|                                                            | **quantiles**  |
|                                                            |                |
|                                                            | **Histograms** |
|                                                            |                |
|                                                            | range          |
+------------------------------------------------------------+----------------+
| dimension reduction analysis                               | PCA            |
+------------------------------------------------------------+----------------+
| statistical tests                                          |                |
+------------------------------------------------------------+----------------+
| linear regression                                          |                |
+------------------------------------------------------------+----------------+

### descriptive statistics

For a quick overview of the content we printed the headlines of the data sets for example for prism.cl.

```{r head_prism.cl, echo=FALSE}
head(prism.cl)
```

We looked for the data distribution for prism.exp, prism.achilles,prism.cnv and prism which seemed to be not normal distributed. This conclusion was confirmed by a Anderson-Darling normality test (p-value \< 2.2e-16).

```{r distribution_histograms, echo=FALSE}
par(mfrow=c(2,2))
hist(unlist(prism), breaks = 60, main = "Histogram of prism", xlab = "treatment effect values")

hist(unlist(prism.exp), breaks = 60, main = "Histogram of prism.exp", xlab = "transcripts per million")

hist(unlist(prism.achilles), breaks = 60,main = "Histogram of prism.achilles", xlab = "gene knockout-scores")

hist(unlist(prism.cnv), main= "Histogram of prism.cnv", xlab="copy number variation", breaks=60)
```

```{r check_normal_distribution, include=FALSE}
#values as vectors
prism_vector <- unlist(prism)
prism.achilles_vector<-unlist(prism.achilles)
prism.exp_vector<- unlist(prism.exp)
prism.cnv_vector<-unlist(prism.cnv)

# Perform the Anderson-Darling normality test
library(nortest)
hyp_test_prism_distribution<- ad.test(prism_vector)
hyp_test_prism_distribution

hyp_test_prism.achilles_distribution<-ad.test(prism.achilles_vector)
hyp_test_prism.achilles_distribution

hyp_test_prism.exp_vector<-ad.test(prism.exp_vector)
hyp_test_prism.exp_vector

hyp_test_prism.cnv_vector<-ad.test(prism.cnv_vector)
hyp_test_prism.cnv_vector
```

The boxplot of the data shows small inter quartile ranges for prism.cnv and prism.achilles and a bigger range for the prism data set. This means we have more spreaded values for prism than for the other two data sets.

```{r boxplot_prism_cnv_ach, echo=FALSE}
boxplot(
  prism_vector,prism.cnv_vector, prism.achilles_vector,
  outline = FALSE,
  names=c("prism", "prism.cnv", "prism.achilles"))
```

### Do some high effective drugs target essential genes of kidney cancer?

#### Which genes are essential for kidney cancer cell lines?

```{r essential_genes, include=FALSE}
#old~new
#cnv~exp
#exp~ach
#identify missing rows in exp that are present in achilles
diff_rows <- setdiff(rownames(prism.exp.kidney), rownames(prism.achilles.kidney))
diff_rows <- unique(c(diff_rows, setdiff(rownames(prism.achilles.kidney), rownames(prism.exp.kidney))))
#print(diff_rows)

#prism.exp.kidney_corr_cleaned<-prism.exp.kidney
#"ACH-000046" "ACH-000171" "ACH-000313" "ACH-000428" "ACH-000429" "ACH-000457"
#which(rownames(prism.exp.kidney)=="ACH-000457")
#2,4,7,9,10,12
prism.exp.kidney_corr_cleaned<- prism.exp.kidney[-c(2,4,7,9,10,12),]

#identify missing columns that are present in prism.exp.kidney_corr_cleaned but not in achilles
diff_cols <- setdiff(colnames(prism.exp.kidney), colnames(prism.achilles.kidney))
diff_cols <- unique(c(diff_cols, setdiff(colnames(prism.achilles.kidney), colnames(prism.exp.kidney))))
#print(diff_cols)


diff_cols2 <- setdiff(colnames(prism.achilles.kidney), colnames(prism.exp.kidney))
diff_cols2<- unique(c(diff_cols, setdiff(colnames(prism.exp.kidney), colnames(prism.achilles.kidney))))
prism.exp.kidney_corr_cleaned<-prism.exp.kidney_corr_cleaned[-which(colnames(prism.exp.kidney) %in% diff_cols)]
prism.achilles.kidney_corr_cleaned<-prism.achilles.kidney[-which(colnames(prism.achilles.kidney) %in% diff_cols2)]

#reorder rows in an ascending way
#1.step:rownames(prism.exp.kidney_corr_cleaned)
prism.exp.kidney_corr_cleaned<- prism.exp.kidney_corr_cleaned[order(rownames(prism.exp.kidney_corr_cleaned)),]
prism.achilles.kidney_corr_cleaned<- prism.achilles.kidney_corr_cleaned[order(rownames(prism.achilles.kidney_corr_cleaned)),]

#reorder columns by names
prism.exp.kidney_corr_cleaned= prism.exp.kidney_corr_cleaned[,colnames(prism.achilles.kidney_corr_cleaned)]

correlation_ach_exp_kidney<-matrix()
for (g in colnames(prism.achilles.kidney_corr_cleaned)) {  
  valse_g_ach_exp_kidney = prism.achilles.kidney_corr_cleaned[,colnames(prism.achilles.kidney_corr_cleaned) == g]  
  valsc_g_ach_exp_kidney = prism.exp.kidney_corr_cleaned[,colnames(prism.exp.kidney_corr_cleaned) == g]  
  correlation_ach_exp_kidney[[g]]=cor(valse_g_ach_exp_kidney, valsc_g_ach_exp_kidney, method= c("spearman"))  
}  
final_correlation_ach_exp_kidney = t(data.frame(correlation_ach_exp_kidney))

colnames_pos_corr <- colnames(final_correlation_ach_exp_kidney)[which(final_correlation_ach_exp_kidney >= quantile(final_correlation_ach_exp_kidney, 0.98, na.rm = TRUE))]

colnames_neg_corr <- colnames(final_correlation_ach_exp_kidney)[which(final_correlation_ach_exp_kidney <= quantile(final_correlation_ach_exp_kidney, 0.02, na.rm = TRUE))]

high_corr_genenames<-c(colnames_neg_corr, colnames_pos_corr)
```

```{r correlation_histogram, echo=FALSE}
hist(correlation_ach_exp_kidney, breaks = 40)
abline(v=c(-0.9,0.9))
```

#### Which drugs do have a high negative effect on the cell lines?

```{r neg_effect_drugs, include=FALSE}
#Set threshold depending on the distribution of prism 
prism.kidney_threshold<- quantile(prism.kidney, 0.02, na.rm=TRUE)

                                                
# Create an empty vector to store the column names
high_effect_drugs_kidney <- c()

# Iterate over each column of the dataset
for (col in colnames(prism.kidney_NA_0)) {
  # Check if any value in the column is lower than the threshold
  if (any(prism.kidney_NA_0[, col] <= prism.kidney_threshold)) {
    # If true, append the column name to the vector
    high_effect_drugs_kidney <- c(high_effect_drugs_kidney, col)
  }
}
#subset of prism.treat for drugs with inhibiting impact on kidney cancer cells
#Show/store the filtered rows
prism.treat.kidney.high_effect_drugs<-print(prism.treat[rownames(prism.treat) %in% high_effect_drugs_kidney,])


prism.treat.kidney.high_effect_drugs$target <- as.character(prism.treat.kidney.high_effect_drugs$target)

##Create a vector with all targets of the high_effect drugs

# Extract all targets as a single character vector
targetgenes_high_effect_drugs_kidney <- 
  unlist(strsplit(prism.treat.kidney.high_effect_drugs$target, ", "))

# Remove any leading/trailing white spaces and NA values
targetgenes_high_effect_drugs_kidneys <- trimws(targetgenes_high_effect_drugs_kidney)
targetgenes_high_effect_drugs_kidney <- targetgenes_high_effect_drugs_kidney[!is.na(targetgenes_high_effect_drugs_kidney)]

# Print the resulting vector of targets
#print(targetgenes_high_effect_drugs_kidney)
```

```{r venn_genes, echo=FALSE}
# Create a list of the gene names in each vector
set1 <- unique(high_corr_genenames)
set2 <- unique(targetgenes_high_effect_drugs_kidney)

list_venn1<-list(set1, set2)

# Create the Venn diagram
library(VennDiagram)
venn1<-venn.diagram(
  list_venn1 ,
  category.names = c("essential genes(high corr. btw. prism.ach & prism.exp)", "target genes of high effective drugs "),
  fill = c("lightblue", "steelblue4"),
  main ="venn diagram",
  main.fontface= "bold",
  sub = "essential genes vs.target genes of high effective drugs ",
  col= "white",
  cat.pos= 9,
  filename = NULL
)
grid.newpage()
grid.draw(venn1)
```

These genes are essential genes as well as targets of high inhibiting drugs:

```{r intersect_genes, echo=FALSE}
#find the genes in the intersection
intr_targets <- intersect(set1, set2)
```

#### What about mutations of those target genes?

```{r prism.snv_subset_targets, include=FALSE}
prism.snv_intr_targets<-print(prism.snv[prism.snv$Hugo_Symbol %in% intr_targets,])
```

```{r}
#count_mutations<-as.vector(prism.snv_intr_targets$Hugo_Symbol)
#count_mutations<-table(vector_count_mutations)

#df.count_mutations<-as.data.frame(count_mutations)

library(ggplot2)
 
##backup-plot 
#ggplot(df.count_mutations, aes(x = vector_count_mutations, y = Freq)) +
 # geom_bar(stat = "identity") +
  #xlab("Gene") +
  #ylab("Number of Mutations") +
  #ggtitle("Number of Mutations per Gene")+
  #theme(axis.text.x = element_text(angle = 90, hjust = 1))

library(RColorBrewer)

# Create a custom color palette based on unique variant classifications
unique_variants <- unique(prism.snv_intr_targets$Variant_Classification)

#color_palette <- terrain.colors(length(unique_variants))
color_palette <- brewer.pal(9, "GnBu")
#RdYlBu


# Plot the histogram  (bars filled with different colors)
ggplot(prism.snv_intr_targets, aes(x = Hugo_Symbol, fill = Variant_Classification)) +
  geom_bar() +
  scale_fill_manual(values = color_palette) +
  labs(x = "gene", y = "Count", title = "Variant classifications of possible target genes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 55, hjust = 1))
```

### Identification of clusters in highly variant drugs

```{r}
#run k-means for k=5,50,100,300,500,700,1000 clusters and check the WSS value for each k. How does WSS evolve with increasing k?
wss_list<-list()
#calculate within-square-distance
for(cancer_type in names(prism.var_list)){
  wss = sapply(c(5,50,100,300,500,700,1000),function(k) { 
    kmeans(x=t(prism.topvar.NA_0), centers =k)$tot.withinss
})
  wss_list[[cancer_type]]<-wss
}
plot(c(5,50,100,300,500,700,1000),wss_list[["Kidney Cancer"]],type='b',pch=19,xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")

##Zoom-In on the elbow-area
wss_list1<-list()
#calculate within-square-distance
for(cancer_type in names(prism.var_list)){
  wss1 = sapply(35:100,function(k) { 
    kmeans(x=t(prism.topvar.NA_0), centers =k)$tot.withinss
})
  wss_list1[[cancer_type]]<-wss1
}
plot(35:100,wss_list1[["Kidney Cancer"]],type='b',pch=19,xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares")

##f체r welchen cancer type ist dieser plot jetzt? ->for-Schleife schreiben und Liste mit den einzelnen wss-zu-k-Werten f체r die einzelnen cancertypes kreieren
```

While there seems to be a knick in the curve at around k=50, zooming-in led to an almost linear curve (here: plot for kidney cancer) with a small knick at k=54, indicating that clustering the highly variant drugs is not meaningful. To confirm this result, closer attention was brought to the content of the clusters, using the clusters in kidney cancer celllines.

```{r}
#assigns each drug/concentration to one of 54 (peak) clusters
km_kidney_cluster = kmeans(t(prism.var_list[["Kidney Cancer"]]), 
            centers = 54, 
            nstart = 10)
print(km_kidney_cluster)

#extracts drug ID s for each cluster and stores them in kidney_cluster_list
kidney_cluster_list<-list()
for(i in 1:54){
  kidney_cluster_list[[i]]<-which(km_kidney_cluster$cluster== i)
}
names(kidney_cluster_list[[25]])

#extracts mode of action for drugs in each cluster and stores them in kidney_cluster_moa list
kidney_cluster_moa<-list()
for(i in 1:54){
  kidney_cluster_moa[[i]]<-prism.treat$moa[rownames(prism.treat) %in% names(kidney_cluster_list[[i]])]
}
print(kidney_cluster_moa[[25]])
```

The extraction of the mode of action and target were repeated for several clusters, but no pattern or logical cluster could be found. This shows that there is not one pathway that is most effective/more represented for several drug/dose combinations in kidney cancer treatment.

### Principal component analysis

#PCA for gene expression

```{r}
pca_exp_celllines=prcomp(prism.exp, center=FALSE, scale=FALSE)
celllines.exp=prism.cl[rownames(prism.cl) %in% rownames(prism.exp),]
# Make another df
toplot.exp = as.data.frame(pca_exp_celllines$x) 
toplot.exp$DepMap_ID = rownames(toplot.exp)

plot_wCells.exp = merge(toplot.exp, celllines.exp, by = "DepMap_ID")

# Plot
ggplot(plot_wCells.exp, 
       aes(x = PC1, y = PC2, color = lineage)) +
  geom_point() 
```

Soft-tissue-celllines and central-nervous-system-celllines seem to cluster together while prostate and kidney cancer are widely spread over the plot. #PCA for copy number variation

```{r}
pca_cnv_celllines=prcomp(prism.cnv, center=FALSE, scale=FALSE)
celllines.cnv=prism.cl[rownames(prism.cl) %in% rownames(prism.cnv),]
# Make another df
toplot.cnv = as.data.frame(pca_cnv_celllines$x) 
toplot.cnv$DepMap_ID = rownames(toplot.cnv)

plot_wCells.cnv = merge(toplot.cnv, celllines.cnv, by = "DepMap_ID")

# Plot
ggplot(plot_wCells.cnv, 
       aes(x = PC1, y = PC2, color = lineage)) +
  geom_point()
```

#PCA for knockout scores

```{r}
achilles_NA_0 <- prism.achilles
achilles_NA_0[is.na(achilles_NA_0)] <- 0 # ersetze alle NA durch 0
pca_achilles_celllines=prcomp(achilles_NA_0, center=FALSE, scale=FALSE)
celllines.achilles=prism.cl[rownames(prism.cl) %in% rownames(prism.achilles),]
# Make another df
toplot.achilles = as.data.frame(pca_achilles_celllines$x) 
toplot.achilles$DepMap_ID = rownames(toplot.achilles)

plot_wCells.achilles = merge(toplot.achilles, celllines.achilles, by = "DepMap_ID")

# Plot
ggplot(plot_wCells.achilles, 
       aes(x = PC1, y = PC2, color = lineage)) +
  geom_point()
```

####Is there a correlation between gene expression and copy number variation?

```{r}
cor_g<-matrix()
for (g in colnames(prism.exp_corr_cleaned)) {  
  valse_g = prism.exp_corr_cleaned[,colnames(prism.exp_corr_cleaned) == g]  
  valsc_g = prism.cnv_corr_cleaned[,colnames(prism.cnv_corr_cleaned) == g]  
  cor_g[[g]]=cor(valse_g, valsc_g, method= c("spearman"))  
}  
final_cor = t(data.frame(cor_g))
h= hist(final_cor)
sum(h$counts[(which(h$breaks <= 0))])
sum(h$counts)
sum(h$counts[(which(h$breaks <= 0))])/sum(h$counts)

#correlation values for kidney cell lines
cor_kidney<-matrix()
for (g in colnames(prism.exp_corr_cleaned)) {  
  valse_g_kidney = prism.exp_corr_cleaned[rownames(prism_exp_list[["Kidney Cancer"]]),colnames(prism.exp_corr_cleaned) == g]  
  valsc_g_kidney = prism.cnv_corr_cleaned[rownames(prism_exp_list[["Kidney Cancer"]]),colnames(prism.cnv_corr_cleaned) == g]  
  cor_kidney[[g]]=cor(valse_g_kidney, valsc_g_kidney, method= c("spearman"))  
}  
kidney_cor = t(data.frame(cor_kidney))
hist(kidney_cor)
```

WHile the general correlation values are widely spread between -0.4 and 0.8 with most values lying on the positive side (84%), the correlation values for kidney cancer seem to normally distributed around 0.025 but with more values having a negative correlation value (34.6%)-\> Statistical test

As the sample size is too large for a Shapiro Wilks test, a qqplot plotting a normal distribution with mean=mean(kidney correlation values) against kidney correlation values was performed. It shows a clear line.

```{r}
qqplot(rnorm(18560,mean=mean(kidney_cor, na.rm = TRUE)),kidney_cor)
ks.test(kidney_cor, "pnorm", mean=mean(kidney_cor, na.rm = TRUE))
```

As p-value \< 2.2e-16, the 0-hypothesis must be rejected, so the data is not normally distributed.

## Discussion

kurze Zusammenfassung der Ergebnisse, Diskussion 체ber gewonnen Erkenntnisse, Einschr채nkungen in der Analyse

## Referenzen
